Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-15 paper/main.tex
starting to address pinned items in slack channel

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    completing three 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about lecture 1, 5
    questions about lecture 2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

@@

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(lecture 1, lecture 2, and general physics knowledge) on each of the three
quizzes. Quiz 1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz 2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., lecture 1), and Quiz 3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., lecture 2).

@@

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture 1 are strongly correlated with the average topic weights
from lecture 1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from lecture 2 are
strongly correlated with the average topic weights from lecture 2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with 
their non-matching question sets (lecture 1 video vs.~lecture 2 questions: $r(13) = 
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; lecture 2 video vs.~lecture 1 questions: $r(13) 
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture 1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the lecture 2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture 1 video vs.~lecture 2 questions; lecture 2 video
vs.~lecture 1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz 1 responses, but should increase when
estimated using Quiz 2 or 3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz 2 versus
Quiz 1 ($t(49) = 8.764,~p < 0.001$) and on Quiz 3 versus Quiz 1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz 2 versus 3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz 2 (but before Quiz 3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes 1 and 2, but
should show a ``boost'' on Quiz 3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes 1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz 3 versus 2 ($t(49) =
10.561,~p < 0.001$) and Quiz 3 versus 1 ($t(49) = 8.969,~p < 0.001$).

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left panel). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered ($t(633) =
0.577,~p = 0.564$). After watching the first video, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle panel) exhibited a positive shift relative
to held-out incorrectly answered questions ($t(633) = 3.961,~p < 0.001$). After
watching the second video, estimated knowledge (from the third quiz;
Fig.~\ref{fig:predictions}, right panel) for \textit{all} questions exhibited a
positive shift. However, the increase in estimated knowledge for held-out
correctly answered questions was larger than for held-out incorrectly answered
questions (estimated knowledge for correctly versus incorrectly answered 
Quiz 3 questions: $t(628) = 2.045,~p = 0.041$).

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of each
    lecture are indicated by dotted lines, and the coordinates of each question are indicated by
    dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps follow
    the same general format as the knowledge maps in Panel A, but here the shading at each
    coordinate indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' between the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word (via their relative sizes) reflected by the blend of topics
    represented at the locations of the stars on the maps. The words' colors
    indicate how much each word is weighted, on average, across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (blue) and
    \textit{Birth of Stars} (green) videos, respectively.}

@@

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz 1 responses (Fig.~\ref{fig:knowledge-maps}A,
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz 2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates traversed by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz~2. This localization is non-trivial: the knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture
(see Eqn.~\ref{eqn:rbf-knowledge}). Finally, the knowledge map estimated from Quiz
3 responses shows a second increase in knowledge, localized to the region
surrounding the embedding of the \textit{Birth of Stars} lecture participants
watched immediately prior to taking Quiz 3.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces} (i.e., lecture 1),
15 about the conceptual content of \textit{Birth of Stars} (i.e., lecture 2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions and answer choices may be found in Table~\questions.

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture 1, the second between
lectures 1 and 2, and the third after viewing lecture 2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about lecture 1, 5 questions about
lecture 2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

After fitting a topic model to the two videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of Euclidean distance, correlation, or other geometric measures). In
general, the similarity between different documents' topic vectors may be used
to characterize the similarity in conceptual content between the documents.

>>>>

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    completing three 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about lecture~1, 5
    questions about lecture~2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

@@

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(lecture~1, lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., lecture~2).

@@

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture~1 are strongly correlated with the average topic weights
from lecture~1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from lecture~2 are
strongly correlated with the average topic weights from lecture~2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with 
their non-matching question sets (lecture~1 video vs.~lecture~2 questions: $r(13) = 
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; lecture~2 video vs.~lecture~1 questions: $r(13) 
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture~1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the lecture~2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture~1 video vs.~lecture~2 questions; lecture~2 video
vs.~lecture~1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

@@

Of course, even though the timecourses in
Figure~\ref{fig:knowledge-timeseries}A and C provide detailed
\textit{estimates} about participants' knowlege, those estimates are only
\textit{useful} to the extent that they accurately reflect what participants
actually know. As one sanity check, we anticipated that the knowledge estimates
should show a content-specific ``boost'' in participants' knowledge after
watching each lecture. In other words, if participants learn about each
lecture's content when they watch each lecture, the knowledge estimates should
reflect that. After watching the \textit{Four Fundamental Forces} lecture,
participants should show more knowledge for the content of that lecture than
they had before, and that knowledge should persist for the remainder of the
experiment. Specifically, knowledge about that lecture's content should be
relatively low when estimated using Quiz~1 responses, but should increase when
estimated using Quiz~2 or~3 responses (Fig.~\ref{fig:knowledge-timeseries}B).
Indeed, we found that participants' estimated knowledge about the content of
the \textit{Four Fundamental Forces} was substantially higher on Quiz~2 versus
Quiz~1 ($t(49) = 8.764,~p < 0.001$) and on Quiz~3 versus Quiz~1 ($t(49) =
10.519,~p < 0.001$). We found no reliable differences in estimated knowledge
about that lecture's content on Quiz~2 versus~3 ($t(49) = 0.160,~p = 0.874$).
Similarly, we hypothesized (and subsequently confirmed) that participants
should show more estimated knowledge about the content of the \textit{Birth of
Stars} lecture after (versus before) watching it
(Fig.~\ref{fig:knowledge-timeseries}D). Specifically, since participants
watched that lecture after taking Quiz~2 (but before Quiz~3), we hypothesized
that their knowledge estimates should be relatively low on Quizzes 1 and 2, but
should show a ``boost'' on Quiz~3. Consistent with this prediction, we found no
reliable differences in estimated knowledge about the \textit{Birth of Stars}
lecture content on Quizzes 1 versus 2 ($t(49) = 1.013,~p = 0.316$), but the
estimated knowledge was substantially higher on Quiz~3 versus~2 ($t(49) =
10.561,~p < 0.001$) and Quiz~3 versus 1 ($t(49) = 8.969,~p < 0.001$).

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left panel). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered ($t(633) =
0.577,~p = 0.564$). After watching the first video, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle panel) exhibited a positive shift relative
to held-out incorrectly answered questions ($t(633) = 3.961,~p < 0.001$). After
watching the second video, estimated knowledge (from the third quiz;
Fig.~\ref{fig:predictions}, right panel) for \textit{all} questions exhibited a
positive shift. However, the increase in estimated knowledge for held-out
correctly answered questions was larger than for held-out incorrectly answered
questions (estimated knowledge for correctly versus incorrectly answered 
Quiz~3 questions: $t(628) = 2.045,~p = 0.041$).

@@

    \caption{\textbf{Mapping out the geometry of knowledge and learning.}
    \textbf{A. Average ``knowledge maps'' estimated using each quiz.} Each map
    displays a 2D projection of the estimated knowledge about the content
    reflected by \textit{all} regions of topic space (see
    \nameref{subsec:knowledge-maps}). The topic trajectories of each
    lecture are indicated by dotted lines, and the coordinates of each question are indicated by
    dots. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,
    and~\individualKnowledgeMapsC. \textbf{B. Average ``learning maps''
    estimated between each successive pair of quizzes.} The learning maps follow
    the same general format as the knowledge maps in Panel A, but here the shading at each
    coordinate indicates the \textit{difference} between
    the corresponding coordinates in the indicated \textit{pair} of knowledge
    maps---i.e., how much the estimated knowledge ``changed'' between the two
    quizzes. Each map reflects an average across all participants. For
    individual participants' maps, see
    Figures~\individualLearningMapsA~and~\individualLearningMapsB. \textbf{C.
    Word clouds for sampled points in topic space.} Each word cloud displays
    the relative weights of each word (via their relative sizes) reflected by the blend of topics
    represented at the locations of the stars on the maps. The words' colors
    indicate how much each word is weighted, on average, across all timepoints'
    topic vectors in the \textit{Four Fundamental Forces} (bluer coloring) and
    \textit{Birth of Stars} (greener coloring) videos, respectively.}

@@

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz~1 responses (Fig.~\ref{fig:knowledge-maps}A,
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz~2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates traversed by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz~2. This localization is non-trivial: the knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture
(see Eqn.~\ref{eqn:rbf-knowledge}). Finally, the knowledge map estimated from Quiz
3 responses shows a second increase in knowledge, localized to the region
surrounding the embedding of the \textit{Birth of Stars} lecture participants
watched immediately prior to taking Quiz~3.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces} (i.e., lecture~1),
15 about the conceptual content of \textit{Birth of Stars} (i.e., lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions and answer choices may be found in Table~\questions.

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture~1, the second between
lectures 1 and 2, and the third after viewing lecture~2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about lecture~1, 5 questions about
lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

@@

After fitting a topic model to the two videos' transcripts, we could use the
trained model to transform arbitrary (potentially new) documents into
$k$-dimensional topic vectors. A convenient property of these topic vectors is
that documents that reflect similar blends of topics (i.e., documents that
reflect similar themes, according to the model) will yield similar coordinates
(in terms of correlation, cosine similarity, Kullback-Leibler divergence,
Euclidean distance, or other geometric measures). In general, the similarity
between different documents' topic vectors may be used to characterize the
similarity in conceptual content between the documents.
